# SCS Curve number method

```{r, message=FALSE, echo=FALSE}
library(rvest)
```

```{r, message=FALSE, include=FALSE}
library(tidyverse)
CN_tbl <- rvest::read_html(
  x = "https://www.hec.usace.army.mil/confluence/hmsdocs/hmstrm/cn-tables") |>
  html_element(xpath = '//*[@id="main-content"]/div/div/div[1]/table') |>
  html_table(trim = TRUE, fill = TRUE)

CN_tbl |>
  janitor::clean_names() |>
  mutate(cover_description = str_remove_all(cover_description, pattern = "\\."))

```

The SCS curve number (CN) is a method developed by the USDA in 1995, when is was formerly named Soil Conservation Service, hence the SCS in the name. [^05_cn-1] CN helps with the estimation of runoff at basins where no runoff has been measured. The CN (curve number) ranges from 0 to 100 and is a dimensionless index representing the combined effect of LU/LC, Soil type and hydrological conditions. CN is used to calculate the potential maximum retention capacity $S$

[^05_cn-1]: https://edepot.wur.nl/183157

$$
S = \dfrac{25400}{CN}-254
$$ where the CN = 0 means complete infiltration, CN = 100 no infiltration at all.

The method estimates the precipitation excess $P_e$ as a function of the cumulative precipitation depth, soil cover, land use, and antecedent soil moisture as

$$
P_e = \begin{cases}
          0, \text{ for }\qquad P < I_a\\
          \dfrac{(P-I_a)^2}{P - I_a + S}\qquad  \text{otherwise }        
      \end{cases}
$$

where $P_e$ is accumulated precipitation excess. $P$ is the accumulated precipitation depth, $I_a$ is the initial abstraction (loss) and $S$ is the potential maximum retention

<!-- $$ -->
<!-- \dfrac{F}{S} = \dfrac{Q}{P-I_a} -->
<!-- $$ -->

$$
I_a = rS\approx0.2\cdot S
$$

$$
S = \dfrac{1000}{\mathrm{CN}}-10\:\mathrm{[mm]}
$$ where $S$ is a potential maximum retention after the initial runoff.

$$
Q = \dfrac{(P-I_a)^2}{(P-I_a)+S}
$$

```{r}
LCLU_tbl <- data.frame(LCLU = c("Pasture", "Road", "Legumes"), 
                       AreaFrac = c(54, 20, 26))
LCLU_tbl

weather_db <- data.frame(
  dtm = seq(as.Date("2024-01-01"), as.Date("2024-11-30"), "1 day"), 
  P = sample(size = 335, c(0, 1), replace = TRUE)*rweibull(335, shape = 1))

head(weather_db)
```

```{r}
# Function to calculate direct runoff using SCS CN method
scs_cn_method <- function(P, CN) {
  
  S <- (25400 / CN) - 254 #<1> 
  
  Ia <- 0.1 * S #<1> 
  
  ifelse(P <= Ia, 0, ((P - Ia)^2) / (P - Ia + S)) #<3> 
}
```

1.  Maximum potential retention (mm)
2.  Initial abstraction/losses (mm)
3.  Calculate runoff

```{r, fig.align='center', fig.width=10}
precipitation <- seq(0, 200, by = 1)  # <1>

curve_numbers <- seq(from = 40, to = 100, by = 5) # <2>

# Calculate runoff for each CN
runoff_results <- sapply(curve_numbers, function(CN) {
  sapply(precipitation, scs_cn_method, CN = CN)
})

# Plot results
plot(NULL, 
     xlim = c(0, max(precipitation)), 
     ylim = c(0, max(runoff_results)),
     xlab = "Precipitation (mm)", 
     ylab = "Runoff (mm)", 
     main = "SCS Curve Number Method: Runoff vs. Precipitation")

for (i in seq_along(curve_numbers)) { #<5>
  lines(precipitation, 
        runoff_results[, i], 
        col = "black", 
        lty = i, 
        lwd = 1)
}

# Add legend
legend("topright", 
       legend = paste("CN =", curve_numbers), 
       col = "black", 
       lty = i, 
       lwd = 1)

```
5. Add lines for each CN
1.  Theoretical rainfall from 0 to 200 mm
2.  CN values
3.  Compute runoff using the SCS CN method

```{r}
dat01138000 <- read.fwf("data/01138000.dly", 
                        widths = c(8, rep(10, 5))) |> 
  mutate(V1 = as.Date(gsub(V1, 
                           pattern = " ", 
                           replacement = "0"), 
                      format = "%Y%m%d")) 

names(dat01138000) <- c("dtm", "prec", "r", "pet", "tmax", "tmin")

dat01138000[which(dat01138000$prec == -99), "prec"] <- NA

head(dat01138000) 
```

```{r, fig.align='center', fig.width=10}
# Example measured precipitation time series (daily data in mm)
precipitation <- dat01138000$prec[100:1000]

# Define Curve Number
CN <- 75  # Example value for a watershed

# Calculate runoff for each day
runoff <- sapply(precipitation, scs_cn_method, CN = CN)

# Create a time vector for plotting
days <- seq_along(precipitation)

# Plot precipitation and runoff
plot(days, 
     precipitation, 
     type = "h", 
     col = "black", 
     lwd = 0.5, 
     ylim = c(0, max(c(precipitation, runoff), na.rm = TRUE)),
     xlab = "Day", 
     ylab = "Value (mm)", 
     main = "CN based Precipitation and Runoff", 
     lty = 3)
lines(x = days, 
      y = runoff, 
      type = "h", 
      col = "#0088BB", 
      lty = 1, 
      lwd = 1.5)
legend("topright", 
       legend = c("Precipitation", "Runoff"), 
       col = c("black", "#0088BB"), 
       lty = c(3, 1), 
       lwd = c(0.5, 1.5))
```

```{r, fig.align='center', fig.width=9, echo=FALSE}
# library(sf)
# library(rworldxtra)
# library(ggplot2)
# library(tidyterra)
# 
# data(countriesHigh)
# 
# domain <- countriesHigh |> 
#   st_as_sf() |> 
#   filter(NAME == "Czech Rep.")
# 
# sfc <- st_sfc(st_polygon(list(rbind(c(st_bbox(domain)[1], st_bbox(cz)[2]),
#                                    c(st_bbox(domain)[3], st_bbox(domain)[2]), 
#                                    c(st_bbox(domain)[3], st_bbox(domain)[4]), 
#                                    c(st_bbox(domain)[1], st_bbox(domain)[4]),
#                                    c(st_bbox(domain)[1], st_bbox(domain)[2])))), crs = 4326) 
# domain_pts <- st_sample(sfc, 100, type = "random")
# x <- domain_pts |> 
# st_union() |> 
#  st_convex_hull()
# 
# 
# plot(domain)
# domain_pts <- domain_pts |> 
#   st_transform(crs = 5514) |>  
#   st_union() |>
#   st_triangulate()
# 
# domain_tr <- st_cast(st_geometry(domain_pts), to = "MULTIPOINT") |> 
#   st_cast(to = "MULTILINESTRING") |>
#   st_cast(to = "POLYGON") 
# 
# sf_polygons <- st_as_sf(data.frame(id = seq_along(domain_tr)), geometry = domain_tr)
# plot(sf_polygons)
#   # st_voronoi(envelope = x) |> 
#   ggplot() +
#     geom_sf(fill = "#efefef") +
#     geom_sf(data = cz_pts) +
#     theme_minimal(base_size = 16)
```

<!-- Now we have domain, created from using a convex hull of sampled points.  -->

::: callout-tip

## Exercise

In practice we would have more than one CN type in the watershed. Estimate the 
runoff from the watershed using the SCS CN method. Using the following
data.
  

| HRU | Area | CN$_i$    |
|-----|------|-----------|
| 1   | 20   | 70        |
| 2   | 16   | 84        |
| 3   | 64   | 74        |

Compare two approaches to calculate runoff.\

  a) Weighted average CN curve.\
  b) Weighted contribution to discharge (separate contribution, first compute 
  runoff and weight apply weights by fraction of area).
:::